---
title: "Visualization 4: karyoploteR"
output:
  html_document:
    df_print: paged
---


# karyoploteR

[karyoploteR](http://bioconductor.org/packages/karyoploteR/) is an 
R/Bioconductor package to plot data along the genome. It is data agnostic, 
meaning that it does not know anything about you data (it doesn't know bioogy!),
and thus is very flexible, because it makes no assumptions about your data. Oh,
and it's also organism agnostic. It can work with ANY genome, so very useful for
those working with non-model organisms!

It follows a philosophy similar to the base R graphics, where you start 
creating a plot and then add additional graphical elements on top of it.

You can get more information reading [the package vignette](http://bioconductor.org/packages/devel/bioc/vignettes/karyoploteR/inst/doc/karyoploteR.html),
reading the inline documentation with `?function_name` or you can go to the
karyoploteR tutorial page at [https://bernatgel.github.io/karyoploter_tutorial/](https://bernatgel.github.io/karyoploter_tutorial/), where you'll also find examples of somewhat complex plots. If 
everything else fail, ask a question on [biostars](https://www.biostars.org/) 
with the tag "karyoploter" or send me an email to bgel@igtp.cat!


To install the package use BiocManager
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("karyoploteR")
```

and once it's installed, load it with `library()`

```{r}
library(karyoploteR)
```



## Let's get started

To create our first karyoplot we'll call the function `plotKaryotype`.

```{r}
plotKaryotype()
```

 

