---
title: "Visualization 1: Introduction"
output:
  html_document:
    df_print: paged
---

# Introduction

  Visualizing your data will help you. Always.
  
  Summary statistics can be useful and tell you a lot... but really, you
  WANT to plot your data.
  
  This is a new take on a classic example called the 
  [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet).
  
  We will use a set of datasets contained in the 
  [datasauRus](https://cran.r-project.org/web/packages/datasauRus/) package. 
  Most of the following code comes from 
  [its vignette](https://cran.r-project.org/web/packages/datasauRus/vignettes/Datasaurus.html).

  We'll start by loading the needed packages
  
```{r Intro1}
  if(!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
  library(dplyr)
  
  if(!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
  library(ggplot2)
  
  if(!requireNamespace("datasauRus", quietly = TRUE)) install.packages("datasauRus")
  library(datasauRus)

```

Let's take a look at the data

```{r Intro2}
  datasaurus_dozen_wide

```

And the the mean of each column
```{r Intro3}
  colMeans(datasaurus_dozen_wide)
```

Or use dplyr to get a few summary statistics for each dataset

```{r Intro4}
  datasaurus_dozen %>% 
    group_by(dataset) %>% 
    summarize(
      mean_x    = mean(x),
      mean_y    = mean(y),
      std_dev_x = sd(x),
      std_dev_y = sd(y),
      corr_x_y  = cor(x, y)
    )
```

With this we can think that the datasets are mostly the same...

Let's plot them. Start with the first one

```{r Intro5}
  plot(x=datasaurus_dozen_wide$away_x, y=datasaurus_dozen_wide$away_y)
```

or use ggplot to plot them all
```{r Intro6, fig.width=12, fig.height=16}
  ggplot(datasaurus_dozen, aes(x=x, y=y, colour=dataset))+
    geom_point()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol=3)

```


  You can find more information on the process to generate such datasets in 
  [the original reasearch page](https://www.autodeskresearch.com/publications/samestats),
  with nice examples on boxplots too!
  
  

## ggplot2 or R base graphics?

  In R there are different ways of creating graphics, but the two main and 
  most used are R base graphics and ggplot2.
  
  ### ggplot2
  [ggplot2](https://ggplot2.tidyverse.org/) is a system to create graphics
  based on "The Grammar of Graphics". In `ggplot2` the graphics are created 
  declaratively, that is, you have to "explain" how to build your plot 
  (i.e. use this variable for the x position, use that variable for the points 
  color, etc...) and ggplot will take care of creating it. 

  For example, using a standard test dataset with information on cars including
  miles per gallon (mpg), car class, etc, we can easily create a plot like this.
  

```{r}
library(ggplot2)

ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
```

  Plots created with ggplot with default parameters are quite good and you can 
  easily change either changing individual parameters or using themes
  

```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point() + theme_minimal()
```

  ggplot2 is part of the [tidyverse](https://www.tidyverse.org/), a "universe" 
  of packages (dplyr, tibble, purrr...) working together with "tidy" data. 

  Working with the tidyverse packages usually forces you to think on your data 
  and analysis in a very specific way. Creating graphics with ggplot2 is not
  an exception, and it forces you to think about your graphics as a mapping of
  data into graphics aesthetics, geometries, etc. This might work for you 
  or might not, and both ggplot2 and the tidyverse have lovers and haters. 
  In any case it is powerful tool and for many people if the first entrance
  to data analysis with R.
  
  There are quite a few great resources to learn data science with R that 
  heavily use ggplot and the tidyverse, since you can get quite satisfying 
  results with a moderate effort. I would specially recommend 
  [R for Data Science](https://r4ds.had.co.nz/) by Hadley Wickham, the author
  of ggplot2, most of the tidyverse and one of the persons behind Rstudio. 
  You can read the whole book online for free or buy a physical book.
  
### R Base Graphics

  As their name suggests, the R base graphics are the basic graphic system of R
  on top of which everything else is built. 
  
  R base graphics are, by default, more crude than ggplot2. They do not include
  colors unless instructed to, they do not no include a legend by default 
  (although adding one is easy)... They might be seen a lower level (as in 
  closer to the machine) than other systems.
  
  In R base graphics you start by creating a plot, and then add additional 
  graphical elements by calling different graphical functions.
  
  Using the same data as for the ggplot2 example
  
```{r}
mpg <- data.frame(mpg)

plot(x=mpg$displ, y=mpg$hwy)
```

We get a BW plot with just the data, axis and not much more. However, we can 
easily change the axis labels and add color and with 
a few more lines we can add a legend and label a couple of points.

```{r}
mpg$class <- as.factor(mpg$class)
plot(x=mpg$displ, y=mpg$hwy, col=mpg$class, pch=19, ylab = "Highway miles per gallon", xlab="engine displacement (liters)")
legend('topright', legend = levels(mpg$class), col = 1:7, cex = 0.8, pch = 19)
max.hwy <- which.max(mpg$hwy)
text(x=mpg$displ[max.hwy], y=mpg$hwy[max.hwy], labels = mpg$model[max.hwy], pos=4)
min.hwy <- which.min(mpg$hwy)
text(x=mpg$displ[min.hwy], y=mpg$hwy[min.hwy], labels = mpg$model[min.hwy], pos=4)
```
  
  With the R base graphics you can create simple graphics easily. While
  defaults are not as pretty as the ggplot2 counterparts, they
  have a looooot of customization potential.
  
### Additional Resources

* [Comparing ggplot2 and R Base Graphics (Flowindata)](https://flowingdata.com/2016/03/22/comparing-ggplot2-and-r-base-graphics/)
* [Tufte in R (comparison of Tufte style graphics with ggplot2 and R base graphics)](http://motioninsocial.com/tufte/)
* [Why I don't use ggplot2 (SimplyStatistics)](https://simplystatistics.org/2016/02/11/why-i-dont-use-ggplot2/)
* [R graphics - ggplot vs base (nimahejazi)](https://blog.nimahejazi.org/posts/r-graphics-ggplot-vs-base/)
* [ggplot2 cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
* [50 ggplot2 Visualizations (with Full R code)](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)



