---
title: "Visualization 1: Introduction"
output:
  html_document:
    df_print: paged
---

# Introduction

  Visualizing your data will help you. Always.
  
  Summary statistics can be useful and tell you a lot... but really, you
  WANT to plot your data.
  
  This is a new take on a classic example called the 
  [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet).
  
  We will use a set of datasets contained in the 
  [datasauRus](https://cran.r-project.org/web/packages/datasauRus/) package. 
  Most of the following code comes from 
  [its vignette](https://cran.r-project.org/web/packages/datasauRus/vignettes/Datasaurus.html).

  We'll start by loading the needed packages
  
```{r Intro1}
  if(!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
  library(dplyr)
  
  if(!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
  library(ggplot2)
  
  if(!requireNamespace("datasauRus", quietly = TRUE)) install.packages("datasauRus")
  library(datasauRus)

```

Let's take a look at the data

```{r Intro2}
  datasaurus_dozen_wide

```

And the the mean of each column
```{r Intro3}
  colMeans(datasaurus_dozen_wide)
```

Or use dplyr to get a few summary statistics for each dataset

```{r Intro4}
  datasaurus_dozen %>% 
    group_by(dataset) %>% 
    summarize(
      mean_x    = mean(x),
      mean_y    = mean(y),
      std_dev_x = sd(x),
      std_dev_y = sd(y),
      corr_x_y  = cor(x, y)
    )
```

With this we can think that the datasets are mostly the same...

Let's plot them. Start with the first one

```{r Intro5}
  plot(x=datasaurus_dozen_wide$away_x, y=datasaurus_dozen_wide$away_y)
```

or use ggplot to plot them all
```{r Intro6, fig.width=12, fig.height=16}
  ggplot(datasaurus_dozen, aes(x=x, y=y, colour=dataset))+
    geom_point()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol=3)

```


  You can find more information on the process to generate such datasets in 
  [the original reasearch page](https://www.autodeskresearch.com/publications/samestats),
  with nice examples on boxplots too!
  
  

## ggplot2 or R base graphics?

  In R  there are two main ways of creating graphics: R base graphics and ggplot2.
  
  


# Visualization types

- Basic plots
  - scatter
  - bars
  - boxplots
  - lines
  - adding additional elements
  
- Heatmaps
  - pheatmap 
  
  
- Graphs (nodes and edges)
  - Gene set enrichment Miriam
  
- karyoploteR







```{r, TEST_CHUNK}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
