---
title: "Visualization 2: Basic Plots"
output:
  html_document:
    df_print: paged
---

We will see a few types of basic plots and how to create them with R base
graphics system

To create the plots we'll use a simple dataset representing the results of
an experiment measuring two proteins (`P1` and `P2`) in two groups cell types
with two conditions for each type: `treated` and `control`.


```{r}

western <- read.delim("./datasets/western.txt")
western

```

```{r}
table(western$Treated)
table(western$CellType)

table(western$Treated, western$CellType)


```

## Barplots

  Barplots represent single measures as the height of rectangles. They are 
  widely used (both with and without error bars) but they tend to hide most
  of the complexity of the data.
  
  Let's see how to plot the mean levels of one of the proteins, P1, in our
  samples. To do that, we first compute the means and then plot them using 
  [`barplot`](https://www.rdocumentation.org/packages/graphics/versions/3.5.2/topics/barplot). 

```{r}
means <- c(treated=mean(western$P1[western$Treated=="treated"]), control=mean(western$P1[western$Treated=="control"]))
barplot(means)
```

And we can separate also the measures by cell type

```{r}

means <- tapply(western$P1, list(western$CellType, western$Treated), mean)
means
```

and plot them one next to the other with `beside=TRUE`

```{r}
barplot(means, beside=TRUE, col = c("dodgerblue", "skyblue"), main="Levels of P1")
legend("topleft", legend = levels(western$CellType), fill =  c("dodgerblue", "skyblue"))
```

And we can make them horizontal with the `horiz` parameter

```{r}
barplot(means, beside=TRUE, col = c("dodgerblue", "skyblue"), main="Levels of P1", horiz = TRUE)
legend("bottomright", legend = levels(western$CellType), fill =  c("dodgerblue", "skyblue"))
```

#### Additional resources

* [Building Barplots with Error Bars](https://datascienceplus.com/building-barplots-with-error-bars/)
* [Bar Plots (Quick-R)](https://www.statmethods.net/graphs/bar.html)
* [ggplot2 barplots : Quick start guide](http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization)


## Boxplots

  Boxplots show a more information than barplots, since they show the median, 
  the interquartile range and usually single out outliers. To create a 
  boxplot we simply need to call the [`boxplot`](https://www.rdocumentation.org/packages/graphics/versions/3.5.2/topics/boxplot)
  function.

  In this case we do not have to compute the summary statistics ourselves, since
  they will be computed by the function itself.

```{r}
  boxplot(western$P1~western$Treated)
```

Using the same formula notation, we can separate the samples by cell type

```{r}
  boxplot(P1~CellType*Treated, data=western,  col =  c("dodgerblue", "skyblue"),
          ylab="P1 level")
```

With box plots we can see that there are more differences by celltype in the 
treated group than we saw using barplots.

However, when the total number of points allow it, it might make sense to add
the actual data points to the plot. We can doo that with [`stripchart`](https://www.rdocumentation.org/packages/graphics/versions/3.5.2/topics/stripchart).

```{r}
  stripchart(P1~CellType*Treated, data=western)
```

As you can see the defaults are... offputting and not very compatible with the 
boxplots we just created. but changing a couple of parameters the chart improves
quite a bit.

```{r}
 stripchart(P1~CellType*Treated, data=western, vertical=TRUE, pch=19)
```

and we can combine both charts into one using the `add` parameter of stripchart
so it does not start a new plot.

```{r}
  boxplot(P1~CellType*Treated, data=western,  ylab="P1 level")
  stripchart(P1~CellType*Treated, data=western, vertical=TRUE,
             add=TRUE, pch=19, col=c("dodgerblue", "skyblue"))
```

### Additional Resources

* [Box Plots (Quick-R)](https://www.statmethods.net/graphs/boxplot.html)
* [ggplot2 box plot : Quick start guide](http://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization)
* [Boxplots with ggplot2](http://t-redactyl.io/blog/2016/04/creating-plots-in-r-using-ggplot2-part-10-boxplots.html)

Violin plots
* [The vioplot package](https://cran.r-project.org/web/packages/vioplot/index.html)รง
* [Split violin plots](https://cran.r-project.org/web/packages/vioplot/vignettes/violin_split.html)
* [Violin plots with ggplot2](http://www.sthda.com/english/wiki/ggplot2-violin-plot-quick-start-guide-r-software-and-data-visualization)


# Visualization types

- Basic plots
  - scatter
  - bars
  - boxplots
  - lines
  - adding additional elements
  